---
output:
  md_document:
    variant: gfm
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#",
  fig.path = "README-"
)
set.seed(0)
#knitr::opts_knit$set(root.dir = 'inst/examples')
```

# deps

> Dependency Management with 'roxygen'-Style Comments

Manage your source code dependencies by decorating your existing R code with special, 'roxygen'-style comments.

[![Build status](https://github.com/analythium/deps/actions/workflows/check.yml/badge.svg)](https://github.com/analythium/deps/actions)

```R
#' @repo deps https://analythium.r-universe.dev
install.packages("deps", repos = "https://analythium.r-universe.dev")

#' @remote analythium/deps
remotes::install_github("analythium/deps")
```

## Why this package?

There are many similar packages out there, some aimed at reproducibility
([packrat](https://CRAN.R-project.org/package=packrat),
[Require](https://CRAN.R-project.org/package=Require),
[switchr](https://cran.r-project.org/package=switchr),
[versions](https://CRAN.R-project.org/package=versions),
[renv](https://CRAN.R-project.org/package=renv),
[capsule](https://github.com/MilesMcBain/capsule)).
Others are focused on dependency management
([remotes](https://CRAN.R-project.org/package=remotes),
[pak](https://CRAN.R-project.org/package=pak)).

**So why do we need another one?**

Full reproducibility is an important and heavyweight aspiration that either needs locally cached package libraries, or an accurate snapshot mirroring the local system (i.e. exactly where the package was installed from). Full reproducibility is often required for reports, markdown based documents, scripts. _A loosely defined project that is combined with strict versioning requirements, often erring on the side of "more dependencies are safer"._

As opposed to this, package-based development is the main use case for dependency management oriented packages. In this case, exact versions are only managed to the extent of avoiding breaking changes (given that testing can surface these). _A package focused workflow combined with a "no breaking changes" philosophy to version requirements, leading to leaner installation._

What if you wanted to combine the best of both approaches? _A loosely defined project with just strict-enough versioning requirements_ -- and all this without having to write a `DESCRIPTION` file. Because why would you need a `DESCRIPTION` file when you have no package? Also, a `DESCRIPTION` file won't let you to pin an exact the package version or to specify alternate CRAN-like repositories...

The answer is deps: **you add comments to your code, deps does the rest**.

```{r}
#' @remote analythium/rconfig@CRAN-v0.1.3
rconfig::config()

#' @repo sf https://r-spatial.r-universe.dev
library(sf)

#' @ver rgl 0.108.3
library(rgl)
```

Once you decorated your code, you can call `deps::create()` to write your `dependencies.json` file.

Use `deps::install()` to install dependencies based on scanning a project folder or using and existing `dependencies.json` file.

## What deps does

Required packages are found using the [`renv::dependencies()`](https://rstudio.github.io/renv/reference/dependencies.html) function. System dependencies are based on [sysreqs.r-hub.io](https://sysreqs.r-hub.io) with a focus on DEB (Debian/Ubuntu) packages (but can be set to RPM).

The packages list found by `renv::dependencies()` is refined and modified by 'roxygen'-style comments. But the packages need to be declared/used somewhere in the source code for the comments to take effect.

If no comment is provided for a given package, its source is assumed to be a CRAN repository, and its version to be the latest given the repositories used at the time of installation.

## Tags

Tags are part of the ['roxygen'-style comments](https://cran.r-project.org/package=roxygen2):

```{r}
#' @<tag> <parameter>
```

where `#'` is followed by space, the tag starting with `@`, space, and some parameters.

deps implements the following tags:

| Tag   | Description | Usage |
|-------|-------|---------|
| `@sys`  | System requirement(s) | `@sys req1,req2,...` |
| `@remote`  | Remote source(s) | `@remote remote1,remote2,...` |
| `@local`  | Local source(s) | `@local path1,path2,...` |
| `@ver`  | Versioned package | `@ver pkg version` |
| `@dev`  | Development package(s) | `@dev pkg1,pkg2,...` |
| `@repo`  | CRAN-like source | `@repo pkg repo` |
| `@repos`  | Global CRAN-like repo(s) | `@repos repo1,repo2,...` |
| `@rver`  | R version | `@rver 4.1.3` |

#### System requirements

Known system requirements can be declared using the `@sys` tag, packages can be separated by commas:

```{r}
#' @sys curl,git
```

These packages are added to the list of requirements identified by the [sysreqs.r-hub.io](https://sysreqs.r-hub.io) database for the non-development packages. The platform (e.g. the default `"DEB"` or `"RPM`) can be set via the `R_DEPS_PLATFORM` environment variable.

#### Remote sources

Remotely sourced packages are declared using the `@remote`tag following the [`remotes`](https://CRAN.R-project.org/package=remotes/vignettes/dependencies.html) specifications:

```{r}
#' @remote analythium/rconfig@CRAN-v0.1.3
rconfig::config()
```

This is effectively a version pinning for remotely sourced packages.

#### Local packages

The `@local` tag can be used to provide a path to local directory, or compressed file (tar, zip, tar.gz, tar.bz2, tgz2, or tbz) to install packages from:

```{r}
#' @local mypackage_0.1.0.tar.gz
library(mypackage)
```

This is effectively the same as:

```{r}
#' @remote local::mypackage_0.1.0.tar.gz
library(mypackage)
```

#### Development packages

Dev packages following the `@dev` decorator are excluded from installation, but are not removed if already installed:

```{r}
#' @dev devtools,roxygen2
requireNamespace("devtools")
```

#### Alternative repos

Use the `@repo` tag to specify an alternative CRAN-like repository, e.g. from the [r-universe](https://r-universe.dev/):

```{r}
#' @repo intrval https://psolymos.r-universe.dev
library(intrval)
```

Use the `@repos` tag to set CRAN-like repositories for all package installation:

```{r}
#' @repos http://cran.r-project.org,https://psolymos.r-universe.dev
```

Use the `@repos` tag to set an [MRAN](https://mran.microsoft.com/documents/rro/reproducibility) checkpoint:

``` r
#' @repos https://mran.microsoft.com/snapshot/2020-01-01
```

#### Versioned packages

When CRAN packages require specific a version, use the `@ver` tag according to the [`remotes::install_version()`](https://remotes.r-lib.org/reference/install_version.html) function specifications:

```{r}
#' @ver mefa4 >= 0.3-0
library(mefa4)
```

#### R version

R version declared using the `@rver` tag is recorded, but this information is not used during installation. The R version can be a major (`3`), a major-minor (`3.6`) or a major-minor-patch (`3.6.3`) version. When the version is partially specified, the highest minor/patch version will be used. Use dots as separators.

## Usage

The `create()` function will crawl the project directory for package dependencies. It will amend the dependency list and package sources based on the comments and query system requirements for the packages where those requirements are known for a particular platform. The summary is written into the `dependencies.json` file. Optionally, the system requirements are written into the `dependencies.json` file as well.

`install()` will look for the `dependencies.json` file in the root of the  project directory and perform dependency installation if the file exists. If the file does not exist, it uses `create()` to create that file before attempting installation but removes the `dependencies.json` file afterwards, leaving no trace.

## deps-cli

A CLI example is given in [`inst/examples/02-cli`](./inst/examples/02-cli/):

```bash
cp inst/examples/03-cli/deps-cli.R /usr/local/bin/deps-cli
chmod +x /usr/local/bin/deps-cli
```

Usage of the CLI as explained in `deps-cli help`:

```
🚀 Quickly install R package dependencies on the command line

👉 MIT (c) Analythium Solutions Inc. 2022-2023
          _                           _ _ 
       __| | ___ _ __  ___        ___| (_)
      / _` |/ _ \ '_ \/ __|_____ / __| | |
     | (_| |  __/ |_) \__ \_____| (__| | |
      \__,_|\___| .__/|___/      \___|_|_|
                |_|                       

🔗 See https://github.com/analythium/deps

deps-cli is based on deps 0.0.2 2022-10-24 

Usage: deps-cli <command> [options]

Commands:
  deps-cli help         Print usage and exit
  deps-cli version      Print version and exit
  deps-cli create       Scan DIR and write dependencies.json
  deps-cli sysreqs      Install system requirements
  deps-cli install      Install R package dependencies
  deps-cli all          create & sysreqs & install in one go

Options:
  --dir DIR             Directory to scan, defaults to .
  --upgrade             Upgrade package dependencies
  --silent              Silent, no info printed

Examples:
  deps-cli help
  deps-cli version
  deps-cli create
  deps-cli create --silent
  deps-cli sysreqs
  deps-cli install --dir /root/app
  deps-cli all --dir /root/app --upgrade
```

Analyze dependencies, install system and R dependencies in 1 line:

```bash
deps all

# same as
deps-cli create && deps-cli sysreqs && deps-cli install
```

`deps-cli install` looks for the following files before attempting to detect dependencies: `renv.lock`, `pkg.lock`, and `DESCRIPTION`.

In a Dockerfile you can:

```Dockerfile
FROM eddelbuettel/r2u:22.04
ENV R_DEPS_PLATFORM="DEB"
RUN R -q -e "install.packages(c('rconfig', 'deps', 'remotes', 'pak', 'renv'), repos = c('https://cloud.r-project.org', 'https://analythium.r-universe.dev'))"
RUN cp -p $(R RHOME)/library/deps/examples/03-cli/deps-cli.R /usr/local/bin/deps-cli
RUN chmod +x /usr/local/bin/deps-cli

COPY ... # your files

RUN deps-cli all
...
```

Use the `R_DEPS_PLATFORM` environment variable to define system dependencies (e.g. `"RPM"`) if not using an image where those are automatically resolved (like the `eddelbuettel/r2u`).

You can simply use the `ghcr.io/analythium/deps:latest` as your parent image where the `deps-cli` is already installed:

```Dockerfile
FROM ghcr.io/analythium/deps:latest
...
```

## Examples

See the [`inst/examples`](./inst/examples/) folder for more examples.

## Notes

The deps package uses a 5 minutes timeout for downloads instead of the default 1 minute (`getOption("timeout")`). When the timeout option or the `R_DEFAULT_INTERNET_TIMEOUT` environment variable is set to a >5 minutes value, it will be respected and timeout will be set to the maximum of the three possible values.

The platform for system requirements (e.g. the default `"DEB"` or `"RPM`) can be set via the `R_DEPS_PLATFORM` environment variable. The default value is `"DEB"`.

The `create()` function prompts the user asking confirmation before writing the `dependencies.json` file. Use `create(ask = FALSE)` to bypass the prompt.

## License

[MIT License](./LICENSE) © 2022-2023 Peter Solymos and Analythium Solutions Inc.
